Human capital drives natural selection in contemporary humans

### 2021-04-21 why-natural-selection.Rmd
Remove the one-period model results.

Here $a$ captures the strength of preference for children. $u(\cdot)$
is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

The *substitution effect* is negative and reflects that when wages increase,
time devoted to childcare costs more in foregone income. The positive *income
effect* depends on the curvature of the utility function, and reflects that when
income is higher, the marginal loss of income from children is less painful.
Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we solve for optimal fertility $N^{*}$, and examine the
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For moderate levels
of risk aversion ($0.5<\sigma<1$):

\begin{enumerate}
\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
\item The relationship is weaker (closer to zero) at higher wages and/or
levels of human capital.
\item The relationship is more negative when the time burden of children,
$b$, is larger.
\end{enumerate}

@@

Instead of identifying human capital $h$ with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education increases period 2 wages, and is
complementary to human capital $h > 0$. For period 2 wages we assume
the simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Maintaining the assumption of CRRA utility, we examine total
fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$. For $0.5 < \sigma < 1$,
facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma = 1$).
In addition, for $\sigma$ in a neighborhood of 1:

\begin{enumerate}
\item [4.]\setcounter{enumi}{4}The fertility-human capital relationship
is weaker at higher levels of education $s$.
\item The relationship is weaker among those who start fertility in period
2 ($N_{1}^{*} = 0$) than among those who start fertility in period
1 ($N_{1}^{*} > 0$).
\end{enumerate}

Facts 1-5 match our empirical results. In this model, polygenic scores which
correlate with human capital would have a negative correlation with fertility at
low income and education levels, and a weaker or zero correlation at higher
levels (Figures \@ref(fig:plot-income-educ-level)a and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
correlated with human capital, the correlation might also be weaker at higher
levels of the score itself, giving rise to diversifying selection, as we observe
for scores for educational attainment (Appendix
Figure \@ref(fig:plot-purifying)). The correlation would be more negative when
the time burden of children $b$ is high, e.g. for single parents (Figure
\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
starting fertility later (Figure \@ref(fig:plot-age-flb)).

Our results are actually stronger than this, in that correlations with
fertility are reversed at higher AFLB. This could be accommodated in the model 
if children have a money cost as well as a time cost, since money costs are less
burdensome at higher wages, i.e. in period 2 for those with high human capital. Appendix Figure \@ref(fig:N-plot-with-m) shows an example.

@@

```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score.", fig.height = 4}

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their effect
on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with two measures of
human capital: earnings in a respondent's first job, and educational attainment.
The relationships are strongly negative. Thus, human capital appears to be relevant
to natural selection. Substitution effects dominate income effects overall, which fits the known association between income and lower fertility 
[@becker1960economic;@jones2006economic]. The correlations reverse when we control
for age at first live birth, suggesting that within AFLB categories, the income effect dominates. (See also the Supplementary Animations for the
uncorrected effects per AFLB group, which show the correlation reversing after
age 22.)

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note]

[^note]: Indeed, in the one-period model above, while $\frac{dN^{*}}{da} > 0$,
this relationship becomes stronger, not weaker, at higher wages.

Another alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be partially captured by our risk-aversion 
parameter $\sigma$, which affects fertility by changing the marginal utility of 
income, but a more direct channel is risky sexual behaviour
[@Mills2020.05.06.081273]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
(younger) age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. This non-result could simply
reflect the imprecision of the risk aversion measure, which is a single yes/no
question. However, the risk aversion measure *does* predict the overall number
of children, highly significantly ($p < 2 \times 10^{-16}$, `r n_risk_sig`/33
regressions). Given that, and the statistical power we get from our sample size, 
we believe that the non-result is real: while risk aversion does predict fertility 
in the sample, it is not an important channel for natural selection.

>>>>

Here $a$ captures the strength of preference for children. $u(\cdot)$
is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

The *substitution effect* is negative and reflects that when wages increase,
time devoted to childcare costs more in foregone income. The positive *income
effect* depends on the curvature of the utility function, and reflects that when
income is higher, the marginal loss of income from children is less painful.

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$,
and in a neighbourhood of $\sigma = 1$:

1. The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
2. The relationship is weaker (closer to zero) at higher wages and/or
  levels of human capital.
3. The relationship is more negative when the time burden of children,
  $b$, is larger.
4. The relationship is weaker at higher levels of education $s$.
5. The relationship is weaker among those who start fertility in period
  2 ($N_{1}^{*} = 0$) than among those who start fertility in period
  1 ($N_{1}^{*} > 0$).

Facts 1-5 parallel our empirical results.[^one-period-results] In this model,
polygenic scores that correlate with human capital would have a negative
correlation with fertility at low income and education levels, and a weaker or
zero correlation at higher levels (Figures \@ref(fig:plot-income-educ-level)a
and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
correlated with human capital, the fertility correlation might also be weaker at
higher levels of the score itself, giving rise to diversifying selection, as we
observe for scores for educational attainment (Appendix Figure
\@ref(fig:plot-purifying)). The correlation would be more negative when the time
burden of children $b$ is high, e.g. for single parents (Figure
\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
starting fertility later (Figure \@ref(fig:plot-age-flb)).[^money-costs]

[^one-period-results]: Facts 1-3 also hold in the one-period model.

[^money-costs]: Our results are actually stronger than this, in that
correlations with fertility are reversed at higher AFLB. This prediction can be
accommodated in the model if children have a money cost as well as a time cost. 
Appendix Figure \@ref(fig:N-plot-with-m) shows an example.

@@

```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score."}

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
on each polygenic score against that score's correlation with two measures of
human capital: earnings in a respondent's first job, and educational attainment.
The relationships are strongly negative. Thus, human capital appears to be relevant
to natural selection. Substitution effects dominate income effects overall, which fits the known association between income and lower fertility 
[@becker1960economic;@jones2006economic]. The correlations reverse when we control
for age at first live birth, suggesting that within AFLB categories, the income effect dominates. (See also the Supplementary Animations for the
uncorrected effects per AFLB group, which show the correlation reversing after
age 22.)

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$, `r n_risk_sig`/33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$,
this effect becomes stronger, not weaker, at higher wages.
